#Lab_03 {lab_03}

rob wells, 8/14/2021

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse here

# Load janitor here


```

### Load data

In this data set, each row represents a city or town in Maryland with the 2010 and 2020 figures for: 

-  Total population 
-  White only population
-  Black or African American only population
-  Hispanic only population 
-  Native American only population
-  Asian only population
-  Pacific Islander (including native Hawiian) only population

Background on this data from [Social Explorer](https://www.socialexplorer.com/data/ACS2016_5yr/metadata/?ds=ACS16_5yr&table=B02001) 
```{r}
# Write code to load the Maryland race by city, place for 2010 - 2020 and call it md_race.
# The data is called lab_03.csv, and is in the same folder as lab_03.qmd (this data notebook)

# your answer here

MarylandData <- rio::import("lab_03.csv" , which = "RaceData")

View(MarylandData)

```


## Answer questions

Q1: Which five cities have the largest overall Hispanic population in 2020? Name the five cities that had the largest Hispanic population in 2020. 

A1:
```{r}

hispanic <- MarylandData %>%
  select(place, x2020_hispanic) %>% 
  slice_max(x2020_hispanic, n=5)

# your answer here
City           Population 
Baltimore City 32627
Wheaton        22787
Germantown     21381
Silver Spring  21087
Aspen Hill     18635
The five cities with the highest Hispanic population are Baltimore City, Wheaton, Germantown, Silver Spring and Aspen Hill.

```

Q2: Find the cities with the largest Black population and arrange the results in descending order. 

A2:

```{r}
#write your code here

black<- MarylandData %>% 
  select(place, x2020_black) %>% 
  arrange(desc(x2020_black))

# your answer here
Baltimore City had the largest Black population in 2020 with a population of 375,002 followed by Waldorf and Bowie City.

```

Q3: Create a table called top_hispanic that shows the top 15 communities with the largest Hispanic population in 2020, and also display the 2010 Hispanic population and the total community population for 2020. Look at your results - what stands out to you? Craft a one-paragraph story pitch for your editor. 

A3:

```{r}

hispanic <- MarylandData %>%
  select(place, x2020_hispanic, x2010_hispanic, x2020_total) %>% 
  slice_max(x2020_hispanic, n=15)

# your answer here
In each of the top 15 hispanic populations in 2020 there have been increases in the hispaic population from 2010 to 2020. In particular, Baltimore City has seen an increase of almost 10,000 hispanics. I would take a deeper look at the increase of hispanic populations in Baltimore city versus the increase of the other races in Baltimore City. If the growth is unique or larger than the other races I would investage why that increase exsists. 
```
