Data Analysis Draft
Jacob Hunting


Introduction 

This data notebook looks at the Maryland State MCAP Testing in Prince George’s County Elementary School. It specifically focuses on the 3rd grade version of the Math and ELA (English Language Arts) and the percentage of the students at each school who are proficient in each subject. It will also look at factors surrounding students in elementary schools such as attendance, race and income level to examine the intersections between these factors and proficiency levels. The MCAP testing data was pulled from the Maryland Government’s Website. The rest of the data was pulled from the shared drive. The data has been cleaned and filtered to focus on PG County.
```{r}
options(scipen=999)
library(tidyverse)
library(janitor)
library(rio)
library(dplyr)
library(leaflet)
library(sf)

```
Load Data 
```{r}
attendance <- rio::import("attendance.rds")
pg_county <- rio::import("pg_county.rds")
pg_elementary <- rio::import("pg_elementary_school_locations.rds")
pg_income <- rio::import("pg_tract_income.rds")
md_race <-read.csv("md_race_all_8_13.csv")
md_math <- rio::import("2023_MCAP_MATH_Administrative_Data_Report_Card.xlsx")
md_ela <- rio::import("2023_MCAP_ELA_Administrative_Data_Report_Card.xlsx")
md_crosswalk <-rio::import("maryland_tracts.xls")
geoid <- rio::import("md_counties_geoids.txt")


```
Clean Data
```{r}
pg_math <- md_math %>% 
  filter(LEA_Name == "Prince George's") %>% 
  filter(Assessment == "Mathematics 03") %>% 
  clean_names() 

pg_ela <- md_ela %>% 
  filter(LEA_Name == "Prince George's") %>% 
  filter(Assessment == "English Language Arts 03") %>% 
    clean_names() 


pg_math_ela <- pg_math %>% 
  inner_join(pg_ela, by="school_name") %>% 
  clean_names() %>% 
  rename(math_profic_pct=proficient_pct_x, ela_profic_pct=proficient_pct_y) %>% 
  select(school_name, math_profic_pct, ela_profic_pct) %>% 
  mutate(pg_math_ela1= as.numeric(math_profic_pct)) 
  

pg_income<- pg_income %>% 
st_as_sf()

pg_elementary2 <- pg_elementary %>% 
inner_join(attendance, by="school_name") %>% 
  st_as_sf()

pg_income_elementary <- pg_elementary2 %>%
  st_join(pg_income) 

pg_elementary_income_attendance <- pg_income_elementary %>% 
inner_join(attendance, by="school_name") %>% 
  clean_names() 
```

```{r}
pg_elementary_income_attendance2<-pg_elementary_income_attendance %>%  
as.data.frame() %>% 
rename(tract = geoid) %>% 
  mutate(tract1= as.numeric(tract))


pg_elementary_income_attendance_race_crosswalk <- pg_elementary_income_attendance2 %>% 
inner_join(md_crosswalk, by=c("tract1" = "tract")) %>% 
  rename(attendance_rate = attend_rate_pct_x)
```

Questions 

Question 1
Is there a correlation between race and attendance in PG County elementary schools? 
```{r}
pg_race_attendance <- pg_elementary_income_attendance_race_crosswalk %>% 
  select(school_name, attendance_rate, black, white, asian) %>%
  arrange(desc(attendance_rate))

  

pg_highest_atten <-pg_race_attendance %>%
  arrange(desc(attendance_rate)) %>% 
  slice_max(attendance_rate, n=3)



pg_lowest_atten <-pg_race_attendance %>%
  arrange(attendance_rate) %>% 
  slice_min(attendance_rate, n=3) %>% 
  


pg_race_attendance <- pg_race_attendance %>% 
ggplot(pg_race_attendance, aes(x=attendance_rate))+
  geom_line(aes(y=white), color = "blue") +
  geom_line(aes(y=black), color = "orange") +
  geom_line(aes(y=asian), color = "purple") +
   labs(
    title = "The racial makeup of the PG County Elementary Schools arranged by Race",
    x = "Attendance Rate",
    Y = "Racial Makeup")
```

Question 2
Is there a correlation between median income and attendance in PG County Elementary Schools? 
```{r}
pg_median_income_attendance <- pg_elementary_income_attendance_race_crosswalk %>% 
  select(school_name, attendance_rate, median_household_income) 

pg_median_income_attendance_high <- pg_median_income_attendance %>% 
  arrange(desc(attendance_rate)) %>% 
  slice_max(attendance_rate, n=3)

pg_median_income_attendance_low <- pg_median_income_attendance %>% 
  arrange((attendance_rate)) %>% 
  slice_min(attendance_rate, n=3)


pg_median_income_attendance %>% 
ggplot(aes(x=attendance_rate, y=median_household_income))+
geom_point()+
labs(x = "Attendance Percentage" , y = "Household Median Income", Title = "Math")


```
Question 3
Is there a correlation between test scores and median household income in PG County Elementary Schools?
```{r}
pg_math_ela_avg <- pg_math %>% 
  inner_join(pg_ela, by="school_name") %>% 
  clean_names() %>% 
  rename(math_profic_pct=proficient_pct_x, ela_profic_pct=proficient_pct_y) %>% 
  select(school_name, math_profic_pct, ela_profic_pct)


#Note to Professor Wells, I wasn't sure the best way to go about using the data that says under 5% I know I could clean it and make is usable for mutates but that would make the data say just 5% or should I omit this from the data entirely. 
```

Question 4
Is there a correlation between test scores and attendance in PG County Elementary Schools? 
```{r}
pg_math_ela_attendence <- pg_math_ela %>% 
  inner_join(attendance, by="school_name") %>% 
  select(school_name, math_profic_pct, ela_profic_pct,attend_rate_pct) 

pg_math_ela_attendence %>% 
ggplot(aes(x=attend_rate_pct, y=math_profic_pct))+
geom_point()+
labs(x = "Attend Percentage" , y = "Percent of Students Proficent in Math", Title = "Math")

 
```

```{r} 
pg_math_ela_attendence <- pg_math_ela_attendence%>% 
ggplot(pg_math_ela_attendence, aes(x=attendance)) +
  geom_point(aes(y= math_profic_pct), color = "red") +
  geom_point(aes(y=(ela_profic_pct)), color = "blue") +
  theme_classic()+
  labs(title = "Math and Ela Proficent Pergentages by Attendance Percentage",
       x = "Attendance Percentage")
#Note to Professor Wells, I'm getting the same error code with this graph and the line graph I was trying to make for question one. I tried to debug it and find out why it wouldn't work, I think its something with the data frames but it didn't like when I switched them to sf's either.
```


Question 5
Looking at the school zone with the highest test scores where do they rank in terms of attendance and median income?
```{r}
pg_math_ela_combined <- pg_math %>% 
  inner_join(pg_ela, by="school_name") %>% 
  clean_names() %>% 
  rename(math_profic_pct=proficient_pct_x, ela_profic_pct=proficient_pct_y, level_4_pct_math=level_4_pct_x, level_4_pct_ela=level_4_pct_y) %>% 
  select(school_name, level_4_pct_math, level_4_pct_ela)
 
```

Memo
The most interesting finding so far has been the correlation between attendance and math proficiency in the PG County Schools. While I’m still working to get the ELA data on the same chart to compare these two factors fully. It is interesting that in the math data, the correlation between higher proficiency levels and attendance does not appear to be very high. It is a natural assumption that schools with more students attending class and learning would have higher testing scores than schools where student attendance is lower. While there is some positive correlation between these factors the correlation is not very strong. For example, the three schools with the highest attendance on the graph are very spread apart with one placing high one placing in the middle and one being closer to the bottom. Overall it's very interesting because it would be assumed that attendance is one of the most important factors. 