#lab_08
Sean Mussenden, Aug 16, 2022

## About this lab

To complete this lab, you need to:
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings
**Task**: Load janitor and the tidyverse
```{r}
library(tidyverse)
library(janitor)

```
## Load data
```{r}
black_pop <- read_csv("data/maryland_black.csv")
hisp_pop <- read_csv("data/maryland_hispanic.csv")
md_pop <- read_csv("data/maryland_population.csv")
white_population <- read_csv("data/maryland_white.csv")
```

**Task**: Load four tables of U.S. Census data.  Each table has one row per Maryland county with 2020 population information.  Each table has these same two columns:

* geoid - a unique identifier for each county
* name - the county name

The third column in each table represents the total population in one of four demographic categories.

* maryland_population.csv - total_pop - total population 
* maryland_hispanic.csv - hispanic_pop - total Hispanic population
* maryland_white.csv - white_pop - total White alone population
* maryland_black.csv - black_pop - total Black alone population

Some of the information in the tables may need to be cleaned or validated in order to produce the correct answer.  

## Questions

**Q1**  What Maryland county has the largest Hispanic population (by percentage of total population)? 
**A1** 
```{r}
hisp_pop <- hisp_pop %>% 
  mutate(case_when(name == "Montgormery County,Maryland"~"Montgomery County, Maryland", TRUE ~name))

master_pop <- md_pop %>% 
  select(everything()) %>% 
  inner_join(black_pop) %>% 
  inner_join(hisp_pop) %>% 
  inner_join(white_pop)

master_pop <- master_pop %>% 
  select(everything()) %>% 
  mutate(pct_hisp = (hispanic_pop/total_pop)*100) %>% 
  arrange(desc(pct_hisp))
```
Prince George's County has the highest population of Hispanics by percentage of population with 18.79% of the population being Hispanic.

**Q2**  What is the total population of people who identify as white alone in Maryland? 
**A2**  
```{r}
white_pop <- white_population %>% 
  mutate(white_pop=case_when(
    name == "Somerset County, Maryland"~white_pop/1000, TRUE ~ white_pop)) %>% 
summarize(white_pop=sum(white_pop))
```
The total population of people who identify as white alone is 3275048

**Q3** What is the average Black population in all Maryland counties? 
**A3** 

#According to the Census Bureau's report published on April 1,2020,  Baltimore City's population was 585,708 people and 61.2% of the population was Black according to the document, so this is 358,453 people.
#According to the Census Bureau's report published on April 1,2020,  Baltimore County's population was 854,535 people and 31.9% of the population was Black according to the document, so this is 272,596 people.


#Link https://www.census.gov/quickfacts/fact/table/baltimorecitymaryland/RHI225222#RHI225222 

```{r}

black_pop <- black_pop %>% 
  mutate(black_pop = (case_when(
    name == "Baltimore County, Maryland" ~ 272596, 
    name == "Baltimore city, Maryland" ~ 358453,
    TRUE ~ black_pop
)))
mean(black_pop$black_pop)
```
The average Black population in all Maryland counties is 75669.88. 